Red Wine Exploration by Paulo Casaretto
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(GGally)
library(dplyr)
library(memisc)
library(DAAG)
library(Simpsons)
theme_set(theme_minimal(20))
```

```{r echo=FALSE, Load_the_Data}
wine <- read.csv('wineQualityReds.csv')

# Transform quality from an integer to a ordered factor
wine$quality <- factor(wine$quality, ordered = T)

# create a rating varible
wine$rating <- ifelse(wine$quality < 5, 'bad', ifelse(
  wine$quality < 7, 'average', 'good'))
wine$rating <- ordered(wine$rating,
                     levels = c('bad', 'average', 'good'))
```

# Univariate Plots Section
```{r echo=FALSE, Summary}
# Summarize the data set
dim(wine)
names(wine)
str(wine)
summary(wine)
```

```{r echo=FALSE, Rating}
ggplot(data=wine, aes(x=quality)) +
  geom_histogram(binwidth = 1)
ggplot(data=wine, aes(x=rating)) +
  geom_histogram(binwidth = 1)
```

The high concentration of wines in the center region and the lack of outliers
might be a problem for generating a predicting model later on.

```{r echo=FALSE, warning=FALSE, fixed.acidity}
ggplot(data=wine, aes(x=fixed.acidity)) +
  geom_histogram(binwidth = 1)
```

There is a high concentration of wines with fixed.acidity close to 8 (the
median) but there are also some outliers that shift the mean up to 9.2.

```{r echo=FALSE, warning=FALSE, volatile.acidity}
qplot(volatile.acidity, data = wine, binwidth = .01)
```

The distribution appears bimodal at 0.4 and 0.6 with some outliers in the higher
ranges.

```{r echo=FALSE, warning=FALSE, citric.acid}
qplot(citric.acid, data = wine, binwidth = .01)
```

Now this is strange distribution. 8% of wines do not present critic acid at all.
Maybe a problem in the data collection process?

```{r echo=FALSE, warning=FALSE, residual.sugar}
qplot(residual.sugar, data = wine, binwidth = .1)
```

A high concentration of wines around 2.2 (the median) with some outliers along
the higher ranges.

```{r echo=FALSE, warning=FALSE, chlorides}
qplot(chlorides, data = wine, binwidth = .01)
```

We see a similar distribution with chlorides.

```{r echo=FALSE, free.sulfur.dioxide}
qplot(free.sulfur.dioxide, data = wine, binwidth = 1)
```

The distributions peaks at around 7 and from then on resembles a long tailed
distribution with very few wines over 60.

```{r echo=FALSE, total.sulfur.dioxide}
qplot(total.sulfur.dioxide, data = wine, binwidth = 1)
```

As expected, this distribution resembles closely the last one.

```{r echo=FALSE, warning=FALSE, density}
qplot(density, data = wine, binwidth = 0.001)
```

The distribution for density has a very normal appearence.

```{r echo=FALSE, pH}
qplot(pH, data = wine, binwidth = 0.01) +
  scale_x_continuous(breaks=c(2.8,3,3.2,3.4,3.6,3.8,4))
```

pH also looks normally distributed.

```{r echo=FALSE, sulphates}
qplot(sulphates, data = wine, binwidth = .1)
```

For sulphates we see a distribution similar to the ones of residual.sugar and
chlorides.

```{r echo=FALSE, warning=FALSE, alcohol}
qplot(alcohol, data = wine, binwidth = .1)
```

We see the same rapid increase and then long tailed distribution as we saw in
sulfur.dioxide. I wonder if there is a correlation between the variables.

# Univariate Analysis

### What is the structure of your dataset?
There are 1599 observation of wines in the dataset with 12 features . There is
one categorical variable (quality) and the others are numerical variables that
indicate wine physical and chemical properties of the wine.

Other observations:
The median quality is 6, which in the given scale (1-10) is a mediocre wine.
The better wine in the sample has a score of 8, and the worst has a score of 3.
The dataset is not balanced, that is, there are a more average wines than poor
or excelent ones and this might prove challenging when designing a predicting 
algorithm.

### What is/are the main feature(s) of interest in your dataset?
The main feature in the data is quality. I'd like to determine which features
determine the quality of wines.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The variables related to acidity (fixed, volatile, citric.acid and pH) might
explain some of the variance. I suspect the different acid concentrations might
alter the taste of the wine.
Also, residual.sugar dictates how sweet a wine is and might also have an
influence in taste.

### Did you create any new variables from existing variables in the dataset?
I created a rating variable to improve the later visualizations.

### Of the features you investigated, were there any unusual distributions? Did 
you perform any operations on the data to tidy, adjust, or change the form of
the data? If so, why did you do this?

Citric.acid stood out from the other distributions. It had (apart from some
outliers) an retangularly looking distribution which given the wine quality
distribution seems very unexpected.

# Bivariate Plots Section

```{r echo=FALSE, Correlation_Matrix}
# in order to calculate the correlation matrix we need to transform the data
# a little
cor(
  wine %>%
    # first we remove unwanted columns
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      # now we translate quality to a number
      quality = as.numeric(quality)
    )
  )
```
 
Alcohol has negative correlation with density. This is expected as alcohol is
less dense than water.

Volatile.acidity has a positive correlation with pH. This is unexpected as pH
is a direct measure of acidity.
Maybe the effect of a lurking variable?

Residual.sugar does not show correlation with quality.
Free.sulfur.dioxide and total.sulfur.dioxide are highly correlated as expected.

Density has a very strong correlation with fixed.acidity.
The variables that have the strongest correlations to quality are
volatile.acidity and alcohol.
 

```{r echo=FALSE, Quality_vs_fixed.acidity}
ggplot(data=wine, aes(x=quality, y=fixed.acidity)) +
  geom_boxplot()
```

As the correlation table showed, fixed.acidity seems to have little to no effect
on quality.

```{r echo=FALSE, Quality_vs_volatile.acidity}
ggplot(data=wine, aes(y=volatile.acidity, x=quality)) +
  geom_boxplot()
```

volatile.acidity seems to be an unwanted feature is wines.
Quality seems to go up when volatile.acidity goes down.
The higher ranges seem to produce more average and poor wines.

```{r echo=FALSE, Quality_vs_citric.acid}
ggplot(data=wine, aes(x=quality, y=citric.acid)) +
  geom_boxplot()
```

We can see the soft correlation between these two variables.
Better wines tend to have higher concentration of citric acid.

```{r echo=FALSE, Quality_vs_residual.sugar}
ggplot(data=wine, aes(x=quality, y=residual.sugar)) +
  geom_boxplot()
```

Contrary to what I initially expected residual.sugar apparently seems to have
little to no effect on perceived quality.

```{r echo=FALSE, Quality_vs_chlorides}
ggplot(data=wine, aes(y=chlorides, x=quality)) +
  geom_boxplot()
```

Altough weakly correlated, a lower concentration of chlorides seem to produce
better wines.

```{r echo=FALSE, Quality_vs_free.sulfur.dioxide}
ggplot(data=wine, aes(y=free.sulfur.dioxide, x=quality)) +
  geom_boxplot()
```

The ranges are really close to each other but it seems too little sulfur dioxide
 and we get a poor wine, too much and we get an average wine. 

```{r echo=FALSE, Quality_vs_total.sulfur.dioxide}
ggplot(data=wine, aes(y=total.sulfur.dioxide, x=quality)) +
  geom_boxplot()
```

```{r echo=FALSE, free_vs_total.sulfur.dioxide}
ggplot(data=wine, aes(y=total.sulfur.dioxide, x=free.sulfur.dioxide)) +
  geom_point()
```

As a superset of free.sulfur.dioxide there is no surprise to find a very similar
distribution here.

```{r echo=FALSE, Quality_vs_density}
ggplot(data=wine, aes(y=density, x=quality)) +
  geom_boxplot()
```

Better wines tend to have lower densities, but this is probably due to the
alcohol concentration. I wonder if density still has an effect if we hold
alcohol constant.

```{r echo=FALSE, Quality_vs_pH}
ggplot(data=wine, aes(y=pH, x=quality)) +
  geom_boxplot()
```

Altough there is definitely a trend (better wines being more acid) there are
some outliers. I wonder how the distribution of the different acids affects this

```{r echo=FALSE}
ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point() +
  geom_smooth(method="lm")
ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point() +
  scale_x_log10(breaks=seq(5,15,1)) +
  xlab("log10(fixed.acidity)") +
  geom_smooth(method="lm")
ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point() +
  geom_smooth(method="lm")
ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point() +
  scale_x_log10(breaks=seq(.1,1,.1)) +
  xlab("log10(volatile.acidity)") +
  geom_smooth(method="lm")
ggplot(data = wine, aes(x = citric.acid, y = pH)) +
  geom_point() +
  geom_smooth(method="lm")
ggplot(data = subset(wine, citric.acid > 0), aes(x = citric.acid, y = pH)) +
  geom_point() +
  scale_x_log10() +
  xlab("log10(citric.acid)") +
  geom_smooth(method="lm")
```

It is really strange that an acid concentration would have a positive
correlation with pH. Maybe Simpsons Paradox?

```{r include=FALSE, Simpsons}
simpsons <- Simpsons(volatile.acidity, pH, data=wine)
```

```{r echo=FALSE, Simpsons.plot}
plot(simpsons)
```

When we clusterize the data and recalculate the regression coefficients there
is change in sign which indicated that there is in fact a lurking variable
that distorts the overall coefficient, indicating the presence of Simpsons
Paradox.

Because we know pH measures acid concentration using a log scale, it is
not surprise to find stronger correlations between pH the log of the acid
concentrations.
We can investigate how much of the variance in pH these tree acidity variables
can explain using a linear model.

```{r echo=FALSE, warning=FALSE, Linear_Model_Acids}
acidsLM <- lm(pH ~ 
                I(log10(citric.acid)) +
                I(log10(volatile.acidity)) +
                I(log10(fixed.acidity)),
                    data = subset(wine,citric.acid > 0))
summary(acidsLM)
df = data.frame(wine$quality )
df$predictions <- predict(acidsLM, wine)
df$error <- (df$predictions - wine$pH)/wine$pH

ggplot(data=df, aes(x=wine.quality, y=error)) +
  geom_boxplot()
```

It seems the three acidity variables can only explain half the variance in PH.
The mean error is specially bad on poor and on excellent wines. This leads me
to believe that there are other component that affect acidity.

```{r echo=FALSE, Quality_vs_sulphates}
ggplot(data=wine, aes(y=sulphates, x=quality)) +
  geom_boxplot()
```

Interesting. Altough there are many outliers in the medium wines, better
wines seem to have a higher concentration of sulphates.

```{r echo=FALSE, Quality_vs_alcohol}
ggplot(data=wine, aes(y=alcohol, x=quality)) + 
  geom_boxplot()
```

The correlation is clear here. With an increase in alcohol graduation we see an
increase in the concentration of better graded wines.
Given the high number of outliers it seems we cannot rely on alcohol alone to
produce better wines.
Let's try using a simple linear model to investigate.

```{r echo=FALSE, Linear_Model_over_alcohol}
alcoholQualityLM <- lm(as.numeric(quality) ~ alcohol,
                    data = wine)
summary(alcoholQualityLM)
df = data.frame(wine$quality )
df$predictions <- predict(alcoholQualityLM, wine)
df$error <- (df$predictions - as.numeric(wine$quality))/as.numeric(wine$quality)

ggplot(data=df, aes(x=wine.quality, y=error)) +
  geom_boxplot()
```

Based on the R-squared value it seems alcohol alone only explains about 22% of
the variance in quality. We're going to need to look at the other variables to
generate a better model.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the
investigation. How did the feature(s) of interest vary with other features in
the dataset?

Fixed.acidity seems to have little to no effect on quality
  
Quality seems to go up when volatile.acidity goes down.
The higher ranges seem to produce more average and poor wines.
  
Better wines tend to have higher concentration of citric acid.

Contrary to what I initially expected residual.sugar apparently seems to have
little to no effect on perceived quality.

Altough weakly correlated, a lower concentration of chlorides seem to produce
better wines.

Better wines tend to have lower densities.

In terms of pH it seems better wines are more acid but there were
many outliers. Better wines also seem to have a higher concentration of
sulphates.

Alcohol graduation has a strong correlation with quality, but like the linear
model showed us it cannot explain all the variance alone.
We're going to need to look at the other variables to
generate a better model.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

I verified the strong relation between free and total sulfur.dioxide.

I also checked the relation between the acid concentration and pH. Of those,
only volatile.acidity surprised me with a positive coefficient for the linear
model.

### What was the strongest relationship you found?

The relationship between the variables total.sulfur.dioxide and
free.sulfur.dioxide.

# Multivariate Plots Section

### Alcohol and other variables

```{r echo=FALSE, warning=FALSE, density_over_alcohol_vs_quality}
ggplot(data = wine,
       aes(y = density, x = alcohol,
           color = quality)) +
  geom_point()
```

When we hold alcohol constant, there is no evidence that density affects
quality which confirms our earlier suspicion.

```{r echo=FALSE, warning=FALSE, sulphates_vs_alcohol_vs_quality}
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5))
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  facet_wrap(~rating)
```

Interesting! It seems that for wines with high alcohol content, having a 
higher concentration of sulphates produces better wines.

```{r echo=FALSE, warning=FALSE, volatile.acidity_vs_alcohol_vs_quality}
ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point()
ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating)
```

The reverse seems to be true for volatile acidity. Having less acetic acid on
higher concentration of alcohol seems to produce better wines.

```{r echo=FALSE, warning=FALSE, pH_vs_alcohol_vs_quality}
ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point()
ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating)
```

Low pH and high alcohol concentration seem to be a good match.

### Acid exploration

```{r echo=FALSE, fixed.acidity_vs_volatile.acidity_vs_quality}
ggplot(data = wine,
       aes(y = fixed.acidity, x = volatile.acidity,
           color = quality)) +
  geom_point()
```

Almost no variance in the y axis compared to the x axis. Lets try the other
acids.

```{r echo=FALSE, citric.acid_vs_volatile.acidity_vs_quality}
ggplot(data = wine,
       aes(y = citric.acid, x = volatile.acidity,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating)
```

High citric acid and low acetic acid seems like a good combination.

```{r echo=FALSE, citric.acid_vs_fixed.acidity_vs_quality}
ggplot(data = wine,
       aes(y = citric.acid, x = fixed.acidity,
           color = quality)) +
  geom_point()
  cor.test(wine$citric.acid, wine$fixed.acidity) 
```

Altough there seems to a correlation between tartaric acid and citric acid
concentrations, nothing stands out in terms of quality.

### Linear model

```{r echo=FALSE, Linear_Model}
set.seed(87654321)
training_data <- sample_frac(wine, .6)
test_data <- wine[ !wine$X %in% training_data$X, ]
m1 <- lm(as.numeric(quality) ~ alcohol, data = training_data)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + fixed.acidity)
m6 <- update(m2, ~ . + pH)
mtable(m1,m2,m3,m4,m5,m6)
df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")
ggplot(data=df, aes(x=quality,y=error)) +
  geom_point()
```

I did not include pH in the same formula with the acids to avoid colinearity
problems.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

High alcohol contents and high sulphate concentrations combined seem to produce
better wines.

### Were there any interesting or surprising interactions between features?



### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

Yes, I created several models. The most prominent of them was composed of
the variables alcohol, sulphates, and the acid variables.
There are two problems with it. First the low R squared score suggest that
there is missing information to propely predict quality.
Second, both the residuals plot and the cross validation favors average wines.
This is probably a reflection of the high number of average wines in the
training dataset or it could mean that there is missing information that would
help predict the edge cases.
I hope that the next course in the nanodegree will help me generate better
models :) .

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, warning=FALSE, citric.acid.final}
qplot(citric.acid, data = wine, binwidth = .01) +
  xlab("citric acid (g / dm^3)")
```

### Description One
This is a very strange distribution. It does not match what we would expect
from a variable collected in a experimental situation. 

### Plot Two

```{r echo=FALSE, warning=FALSE, sulphates_vs_alcohol_vs_quality.final}
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("sulphates (potassium sulphate - g / dm3)") +
  xlab("alcohol (% by volume)")
```

### Description Two
High alcohol contents and high sulphate concentrations combined seem to
produce better wines.

### Plot Three

```{r echo=FALSE, final_linear_model_residuals}
df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")
ggplot(data=df, aes(x=quality,y=error)) +
  geom_point()
```


### Description Three
The linear model with the highest R squared value could only explain around 35%
of the variance in quality.
Also, the clear correlation showed by the residual plot earlier seems to reinforce that there is missing information to better predict both poor and excellent wines.

------
# Reflection

The wine data set contains information on the chemical properties of a selection
of wines collected in 2009. It also includes sensorial data (wine ranking).

I started by looking at the individual distributions of the variables,
trying to get a feel for each one.

The first thing I noticed was the high concentration of wines in the middle
ranges of the ranking, that is, average tasting wines. This proved to be very
problematic during the analysis as I kept questioning myself wether there was a
true correlation between two variables or it was just a coincidence given the
lack of "outlier" (poor and excellent) wines.

Out of the chemical varibles, the only one that stood out was the concentration
of citric acid (variable name citric.acid). First thing i noticed was the high
number of wines that had no citric.acid at all. My initial thought was a data
collection error, but upon
[researching the subject](https://en.wikipedia.org/wiki/Acids_in_wine#Citric_acid),
I found out that citric acid is sometimes added to wines to boost overall
acidity, so it makes sense that some wines would have none.
Nonetheless this variable also showed a strange distribution with some peaks but
showing an almost rectangular distribution specially in the 0-0.5 range.

All of the other variables showed either an normal or long tailed looking
distribution.

After exploring the individual variables, I proceded to investigate the
relationships between each input variable and the outcome variable quality.

The most promissing varibles were alcohol concentration, sulphates and the
individual acid concentrations.

I also tried investigating the effect of each acid in the overall pH for the
wine. I used scatterplots to explore the relationships graphically and also
generated a linear model to check how much of pH the three variables accounted
for.

The first surprise here was finding that the correlation between acetic
acid concentration and pH was positive. I immediately suspected this was the
result of some lurking variable (Simpsons paradox) and with the help of the
"Simpsons" package I confirmed that suspicion.

The second finding was discovering that the concentration of the three acids
only account for less than half of the variance in pH. I interpreted this as
a sign that there more components affecting acidity that were not measured.

On the final part of the analysis I tried using multivariate plots to
investigate if there were interesting combinations of variables that might
affect quality. I also used a multivariate plot to confirm that density did not
have an effect on quality when holding alcohol concentration constant.

In the end, the produced model could not explain much of the variance in
quality. This is further corroborated acidity analysis.

For future studies, it would be interesting to mesure more acid types in the
analysis. [Wikipedia](https://en.wikipedia.org/wiki/Acids_in_wine) for example,
suggests that malic and lactic acid are important in wine taste and these were
not included in this sample.

Also, I think it would be interesting to include each wine critic judgement as
separate entry in the dataset. After all,  each individual has a different taste
and is subject to prejudice and other distorting factors. I believe that 
having this extra information would add more value to the analysis.